Weighted Model Counting with Conditional Weights

### 2020-07-30 paper/paper.tex
Black-and-white-friendly plots

\begin{itemize}
\item The Main Narrative
  \begin{enumerate}
  \item When weights are defined on literals, the measure on the free BA is
    fully independent.
  \item This means that the BA itself must be larger (i.e., have additional
    `meaningless' literals) to turn any probability distribution into an
    independent one.
  \item We show how we can define conditional weights on literals, allowing us
    to encode any probability distribution into a Boolean algebra that's not
    necessarily independent and thus can be smaller.
  \item We demonstrate a specific example of this by presenting a new way to
    encode Bayesian networks into instances of WMC and adapting a WMC algorithm
    (ADDMC) to run on the new format.
  \item We show that this results in significantly faster inference.
  \item We show that our encoding results in asymptotically fewer literals and
    fewer ADDs, and thus a simpler problem.
  \item (Maybe) we experimentally demonstrate a phase transition based on the
    number of variables per ADD.
  \end{enumerate}
\item Potential criticism may be that this is a step backwards and doesn't allow
  us to use SAT-based techniques for probabilistic inference. However, they can
  still be used for the 'theory+query' part.
  \begin{itemize}
  \item Zero-probability weights and one-probability weights can be interpreted
    as logical clauses. This doesn't affect ADDMC but could be useful for other
    solvers.
  \end{itemize}
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow.
\item
  Algorithms
  \begin{itemize}
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} (rediscovered the
    multiplicativity of BAs in different words) (with optimal settings)
  \item Cachet \cite{DBLP:conf/sat/SangBBKP04}
  \item c2d \cite{DBLP:conf/ecai/Darwiche04}
  \item d4 \cite{DBLP:conf/ijcai/LagniezM17} (closed source, boo!)
  \item miniC2D  \cite{DBLP:conf/ijcai/OztokD15}
  \end{itemize}
\item Notable previous/related work
  \begin{itemize}
  \item Hailperin's approach to probability logic
    \cite{DBLP:journals/ndjfl/Hailperin84}
  \item Nilsson's (somewhat successful) probabilistic logic
    \cite{DBLP:journals/ai/Nilsson86,DBLP:journals/ai/Nilsson93}
  \item Measures on Boolean algebras
    \begin{itemize}
    \item On possibility and probability measures in finite Boolean algebras
      \cite{DBLP:journals/soco/CastineiraCT02}
    \item Representation of conditional probability measures
      \cite{krauss1968representation}
    \end{itemize}
  \end{itemize}
\item Intuitively, a measure is just like a probability, except it's in
  $\mathbb{R}_{\ge 0}$ instead of $[0, 1]$.
\end{itemize}

@@

\begin{definition}[Measures and weight functions] \label{def:wmc}
  Let $U$ be a set.
  \begin{itemize}
  \item A \emph{measure} is a function $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$
    such that $M(\bot) = 0$ and
    \[
      M(a \lor b) = M(a) + M(b)
    \]
    for all $a, b \in 2^{2^U}$ whenever $a \land b = \bot$.
  \item A \emph{weight function} is a function $W\colon 2^U \to \mathbb{R}_{\ge
      0}$.
  \item A weight function is \emph{factored} if $W = \prod_{x \in U} W_x$ for
    some functions $W_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$.
  \item We say that a weight function $W\colon 2^U \to \mathbb{R}_{\ge 0}$
    \emph{induces} a measure $M_W\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if
    \begin{equation} \label{eq:induced_measure}
      M_W(x) = \sum_{\{u\} \le x} W(u).
    \end{equation}
  \item A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is
    \emph{factorable} if there exists a factored weight function $W\colon 2^U
    \to \mathbb{R}_{\ge 0}$ that induces $M$.
  \end{itemize}
\end{definition}

@@

\begin{table}
  \caption{An overview of notation for a logic defined over two atoms. The
    elements in both columns are listed in the same order.}
  \label{tbl:notation_example}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Name in logic & Set-theoretic notation & Boolean-algebraic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\emptyset, \{a\}, \{b\}, \{a, b\}$ & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ \\
    & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ & $\top$ \\
    & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{a\}, \{a, b\} \}$ & $\neg a \lor \neg b, a \to b$ \\
    & $\{ \emptyset, \{b\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ & $b \to a, a \lor b$ \\
    & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ & $\neg b, \neg a, a \leftrightarrow b$ \\
    & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ & $(a \land \neg b) \lor (b \land \neg a), a, b$ \\
    & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\emptyset$ & $\bot$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Given the theory $\Delta$, we can also compute the probability of a query $b$ as
\cite{DBLP:conf/uai/Belle17,DBLP:conf/aaai/SangBK05}
\[
  \Pr_{\Delta, w}(b) = \frac{\WMC(\Delta \land b)}{\WMC(\Delta)}.
\]
The same thing can be accomplished using the algebraic formulation, with $M$
replacing $\mathrm{WMC}$.

@@

\section{Limitations of Factorable Measures}

\todo[inline,caption={}]{
  Examples (must):
  \begin{itemize}
  \item an example of a measure that cannot be expressed using a factored weight
    function,
  \item an example of how to express it using conditional weight functions.
  \end{itemize}
  Also, define `conditional weight function'.
}

@@

\begin{theorem}
  A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is factorable if and only
  if
  \begin{equation} \label{eq:wmccondition}
  M(u \land v) = M(u)M(v)
  \end{equation}
  for all distinct $u, v \in U \cup \{ \neg w \mid w \in U \}$ such that $u \ne
  \neg v$.
\end{theorem}
\begin{proof}
  ($\Leftarrow$) For each $x \in U$, let $W_x\colon 2^{\{x\}} \to
  \mathbb{R}_{\ge 0}$ be defined by $W_x(\{ x \}) = M(x)$ and $W_x(\emptyset) =
  M(\neg x)$. Let $M_W$ be the measure induced by
  \[
    W = \prod_{x \in U} W_x.
  \]
  We will show that $M = M_W$. First, note that $M_w(\bot) = 0 = M(\bot)$ by
  \cref{def:wmc,prop:measure}. Second, let
  \[
    a = \bigwedge_{u \in U} a_u
  \]
  be an atom in $2^{2^U}$ such that $a_u \in \{ u, \neg u \}$ for all $u \in U$.
  Then
  \[
    M_W(a) = W(\{ a_u \mid u \in U \}) = \prod_{u \in U} W_u(\{a_u\}) = \prod_{u
      \in U} M(a_u) = M \left(\bigwedge_{u \in U} a_u \right) = M(a)
  \]

@@

  ($\Rightarrow$) For the other direction, we are given a factored weight
  function
  \[
    W = \prod_{x \in U} W_x,
  \]
  and we want to show that its induced measure $M_W$ satisfies
  \cref{eq:wmccondition}. Let $k_u, k_v \in U \cup \{ \neg w \mid w \in U \}$ be
  such that $k_u \in \{ u, \neg u \}$, $k_v \in \{ v, \neg v \}$, and $u \ne v$.
  We then want to show that
  \begin{equation} \label{eq:to_prove3}
    M_W(k_u \land k_v) = M_W(k_u)M_W(k_v)
  \end{equation}
  which is equivalent to
  \begin{equation} \label{eq:to_prove4}
    M_W(k_u \land k_v) \cdot M_W(\neg k_u \land \neg k_v) = M_W(k_u \land \neg k_v) \cdot M_w(\neg k_u \land k_v)
  \end{equation}
  by \cref{lemma:before_theorem}. Then
  \begin{align*}
    M_W(k_u \land k_v) &= \sum_{\{a\} \le k_u \land k_v} W(a) = \sum_{\{a\} \le k_u \land k_v} \prod_{x \in U} W_x(a) \\
                        &= \sum_{\{a\} \le k_u \land k_v} W_u(a_u)W_v(a_v) \prod_{x \in U \setminus \{ u, v \}} W_x(a) = \sum_{\{a\} \le k_u \land k_v} W_u(k_u)W_v(k_v) \prod_{x \in U \setminus \{ u, v \}} W_x(a) \\
    &= W_u(k_u)W_v(k_v) \sum_{\{a\} \le k_u \land k_v} \prod_{x \in U \setminus \{ u, v \}} W_x(a) = W_u(k_u)W_v(k_v)C,
  \end{align*}
  where
  \[
    C = \sum_{\{a\} \le k_u \land k_v} \prod_{x \in U \setminus \{ u, v \}} W_x(a),
  \]
  and is the same for $M_W(\neg k_u \land k_v)$, $M_W(k_u \land \neg k_v)$, and
  $M_W(\neg k_u \land \neg k_v)$ as well. But then \cref{eq:to_prove4} becomes
  \[
    W_u(k_u)W_v(k_v)W_u(\neg k_u)W_v(\neg k_v)C^2 = W_u(k_u)W_v(\neg
    k_v)W_u(\neg k_u)W_v(k_v)C^2
  \]
  which is trivially true.
\end{proof}

@@

A \emph{Bayesian network} is a directed acyclic graph with random variables as
vertices that defines a probability distribution over them. Let $V$ denote the
set of random variables and $X \in V$ be a random variable. Let $\im X$ denote
the set of possible values of $X$ and $\mathrm{pa}(X)$ denote the set of its
parents. The full probability distribution is then equal to
\begin{equation} \label{eq:factorisation}
  \prod_{X \in V} \Pr(X \mid \mathrm{pa}(X)).
\end{equation}
If all random variables are discrete (and we only consider discrete Bayesian
networks in this paper), the probabilities contained in each factor of
\cref{eq:factorisation} can be summarised in a table known as a
\emph{conditional probability table} (CPT). See \cref{fig:example_bn} for an
example Bayesian network which we will refer to throughout this section. For
this network, $V = \{ W, F, T \}$, $\mathrm{pa}(W) = \emptyset$, $\mathrm{pa}(F)
= \mathrm{pa}(T) = \{ W \}$, $\im W = \im F = \{0, 1 \}$, and $\im T = \{ l, m,
h \}$.

\begin{definition}[Indicator variables]
  Let $X \in V$ be a random variable. If $X$ is binary (i.e., $|\im X| = 2$), we
  can arbitrary identify one of the values as $1$ and the other one as $0$ (i.e,
  $\im X \cong \{ 0, 1 \}$). Then $X$ can be represented by a single
  \emph{indicator variable} $\lambda_{X=1}$. For notational simplicity, for any
  set $S$, we write $\lambda_{X=0} \in S$ or $S = \{ \lambda_{X=0}, \dots \}$ to
  mean $\lambda_{X=1} \not\in S$,

  On the other hand, if $X$ is not binary, we represent $X$ with $|\im X|$
  indicator variables, one for each value. We let
  \[
    E(X) = \begin{cases}
      \{ \lambda_{X=1} \} & \text{if } |\im X| = 2 \\
      \{ \lambda_{X=x} \mid x \in \im X \} & \text{otherwise.}
    \end{cases}
  \]
  denote the set of indicator variables for $X$ and
  \[
    E^*(X) = E(X) \cup \bigcup_{Y \in \mathrm{pa}(X)} E(Y).
  \]
  denote the set of indicator variables for $X$ and its parents in the Bayesian
  network. Finally, let
  \[
    U = \bigcup_{X \in V} E(X)
  \]
  denote the set of all indicator variables for all random variables in the
  Bayesian network.
\end{definition}

@@

\begin{algorithm}
  \caption{Encoding a Bayesian network as a function $2^U \to \mathbb{R}_{\ge
      0}$}
  \label{alg:encoding}
  \KwData{a Bayesian network with vertices $V$ and probability distribution
    $\Pr$}
  \KwResult{a function $\phi\colon 2^U \to \mathbb{R}_{\ge 0}$}
  $\phi \gets 1$\;
  \For{$X \in V$}{
    \textit{let} $\mathrm{pa}(X) = \{ Y_1, \dots, Y_n \}$\;
    $\mathrm{CPT}_X \gets 0$\;
    \eIf{$|\im X| = 2$}{
      \For{$(y_1, \dots, y_n) \in \prod_{i = 1}^n \im Y_i$}{
        $p_1 \gets \Pr(X = 1 \mid Y_1 = y_1, \dots, Y_n = y_n)$\;
        $p_0 \gets \Pr(X \ne 1 \mid Y_1 = y_1, \dots, Y_n = y_n)$\;
        $\mathrm{CPT}_X \gets \mathrm{CPT}_X + p_1[\lambda_{X=1}] \cdot
        \prod_{i=1}^n [\lambda_{Y_i=y_i}] + p_0 \overline{[\lambda_{X=1}]} \cdot
        \prod_{i=1}^n [\lambda_{Y_i=y_i}]$\;
      }
    }{
      \textit{let} $\im X = \{ x_1, \dots, x_m \}$\;
      \For{$x \in \im X$ {\rm \textbf{and}} $(y_1, \dots, y_n) \in \prod_{i =
          1}^n \im Y_i$}{
        $p_x \gets \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n)$\;
        $\mathrm{CPT}_X \gets \mathrm{CPT}_X + p_x[\lambda_{X=x}] \cdot
        \prod_{i=1}^n [\lambda_{Y_i=y_i}] + \overline{[\lambda_{X=1}]} \cdot
        \prod_{i=1}^n [\lambda_{Y_i=y_i}]$\;
      }
      $\mathrm{CPT}_X \gets \mathrm{CPT}_X \cdot \left( \sum_{i=1}^m [\lambda_{X
          = x_i}] \right) \cdot \prod_{i=1}^m \prod_{j=i+1}^m
      (\overline{[\lambda_{X = x_i}]} + \overline{[\lambda_{X = x_j}]})$\;
    }
    $\phi \gets \phi \cdot \mathrm{CPT}_X$\;
  }
  \Return{$\phi$}\;
\end{algorithm}

@@

\begin{lemma} \label{lemma:cpt}
  Let $X \in V$ be a random variable with parents $\mathrm{pa}(X) = \{ Y_1,
  \dots, Y_n \}$. Then $\mathrm{CPT}_X\colon 2^{E^*(X)} \to \mathbb{R}_{\ge 0}$
  is such that for any $x \in \im X$ and $(y_1, \dots, y_n) \in \prod_{i=1}^n
  \im Y_i$,
  \[
    \mathrm{CPT}_X (\{ \lambda_{X=x} \} \cup \{ \lambda_{Y_i=y_i} \mid i = 1,
    \dots, n \}) = \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n).
  \]
\end{lemma}
\begin{proof}
  Let $\tau = \{ \lambda_{X=x} \} \cup \{ \lambda_{Y_i=y_i} \mid i = 1, \dots, n
  \}$. If $X$ is binary, then $\mathrm{CPT}_X$ is a sum of $2\prod_{i=1}^n |\im
  Y_i|$ terms, one for each possible assignment of values to variables $X, Y_1,
  \dots, Y_n$. Exactly one of these terms is nonzero when applied to $\tau$, and
  it is equal to $\Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n)$ by definition.

  If $X$ is not binary, then
  \[
    \left( \sum_{i=1}^m [\lambda_{X = x_i}] \right)(\tau) = 1,
  \]
  and
  \[
    \left( \prod_{i=1}^m \prod_{j=i+1}^m (\overline{[\lambda_{X = x_i}]} +
      \overline{[\lambda_{X = x_j}]}) \right)(\tau) = 1,
  \]
  so, by a similar argument as before,
  \[
    \mathrm{CPT}_X(\tau) = \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n).
  \]
\end{proof}

>>>>

\begin{itemize}
\item The Main Narrative
  \begin{enumerate}
  \item When weights are defined on literals, the measure on the free BA is
    fully independent.
  \item This means that the BA itself must be larger (i.e., have additional
    `meaningless' literals) to turn any probability distribution into an
    independent one.
  \item We show how we can define conditional weights on literals, allowing us
    to encode any probability distribution into a Boolean algebra that's not
    necessarily independent and thus can be smaller.
  \item We demonstrate a specific example of this by presenting a new way to
    encode Bayesian networks into instances of WMC and adapting a WMC algorithm
    (ADDMC) to run on the new format.
  \item We show that this results in significantly faster inference.
  \item We show that our encoding results in asymptotically fewer literals and
    fewer ADDs.
  \end{enumerate}
\item Potential criticism may be that this doesn't allow us to use SAT-based
  techniques for probabilistic inference. However, they can still be used for a
  significant part of the encoding.
  \begin{itemize}
  \item Zero-probability weights and one-probability weights can be interpreted
    as logical clauses. This doesn't affect ADDMC but could be useful for other
    solvers.
  \end{itemize}
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow.
\item
  Algorithms
  \begin{itemize}
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} (rediscovered the
    multiplicativity of BAs in different words) (with optimal settings)
  \item Cachet \cite{DBLP:conf/sat/SangBBKP04}
  \item c2d \cite{DBLP:conf/ecai/Darwiche04}
  \item d4 \cite{DBLP:conf/ijcai/LagniezM17} (closed source, boo!)
  \item miniC2D  \cite{DBLP:conf/ijcai/OztokD15}
  \end{itemize}
\item Notable previous/related work
  \begin{itemize}
  \item Hailperin's approach to probability logic
    \cite{DBLP:journals/ndjfl/Hailperin84}
  \item Nilsson's (somewhat successful) probabilistic logic
    \cite{DBLP:journals/ai/Nilsson86,DBLP:journals/ai/Nilsson93}
  \item Measures on Boolean algebras
    \begin{itemize}
    \item On possibility and probability measures in finite Boolean algebras
      \cite{DBLP:journals/soco/CastineiraCT02}
    \item Representation of conditional probability measures
      \cite{krauss1968representation}
    \end{itemize}
  \end{itemize}
\item Intuitively, a measure is just like a probability, except it's in
  $\mathbb{R}_{\ge 0}$ instead of $[0, 1]$.
\end{itemize}

@@

\begin{definition}[Measures and weight functions] \label{def:wmc}
  Let $U$ be a set.
  \begin{itemize}
  \item A \emph{measure} is a function $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$
    such that $M(\bot) = 0$, and $M(a \lor b) = M(a) + M(b)$ for all $a, b \in
    2^{2^U}$ whenever $a \land b = \bot$.
  \item A \emph{weight function} is a function $W\colon 2^U \to \mathbb{R}_{\ge
      0}$.
  \item A weight function is \emph{factored} if $W = \prod_{x \in U} W_x$ for
    some functions $W_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$.
  \item We say that a weight function $W\colon 2^U \to \mathbb{R}_{\ge 0}$
    \emph{induces} a measure $M_W\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if
    \begin{equation} \label{eq:induced_measure}
      M_W(x) = \sum_{\{u\} \le x} W(u).
    \end{equation}
  \item A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is
    \emph{factorable} if there exists a factored weight function $W\colon 2^U
    \to \mathbb{R}_{\ge 0}$ that induces $M$.
  \end{itemize}
\end{definition}

@@

\begin{table}
  \caption{An overview of notation for a logic defined over two atoms. The
    elements in both columns are listed in the same order.}
  \label{tbl:notation_example}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Name in logic & Boolean-algebraic notation & Set-theoretic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\emptyset, \{a\}, \{b\}, \{a, b\}$ \\
    & $\top$ & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg a \lor \neg b, a \to b$ & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{a\}, \{a, b\} \}$ \\
    & $b \to a, a \lor b$ & $\{ \emptyset, \{b\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg b, \neg a, a \leftrightarrow b$ & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ \\
    & $(a \land \neg b) \lor (b \land \neg a), a, b$ & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ \\
    & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\bot$ & $\emptyset$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Given the theory $\Delta$, we can also compute the probability of a query $b$ as
\cite{DBLP:conf/aaai/SangBK05}
\[
  \Pr_{\Delta, w}(b) = \frac{\WMC(\Delta \land b)}{\WMC(\Delta)}.
\]
The same thing can be accomplished using the algebraic formulation, with $M$
replacing $\mathrm{WMC}$.

@@

\subsection{Limitations of Factorable Measures}

\todo[inline,caption={}]{
  Clarify what is meant by`conditional weight function'.
}

\begin{example}
  Let $U = \{a, b\}$ be a set of atoms and $M\colon 2^{2^U} \to \mathbb{R}_{\ge
    0}$ a measure defined as:\footnote{The value of $M$ on any other element of
    the Boolean algebra can be deduced using the definition.}
  \begin{align*}
    M(a \land b) &= 0.72, \\
    M(a \land \neg b) &= 0.18, \\
    M(\neg a \land b) &= 0.07, \\
    M(\neg a \land \neg b) &= 0.03.
  \end{align*}
  If $M$ could be represented using traditional (factored) WMC, we would have to
  find two weight functions $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and
  $W_b\colon 2^{\{b\}} \to \mathbb{R}_{\ge 0}$ such that $W = W_a \cdot W_b$
  induces $M$, i.e., $W_a$ and $W_b$ would have to satisfy this system of
  equations:
  \begin{align*}
    W_a(\{a\}) \cdot W_b(\{b\}) &= 0.72 \\
    W_a(\{a\}) \cdot W_b(\emptyset) &= 0.18 \\
    W_a(\emptyset) \cdot W_b(\{b\}) &= 0.07 \\
    W_a(\emptyset) \cdot W_b(\emptyset) &= 0.03,
  \end{align*}
  which has no solutions.

  Alternatively, we can let $b$ depend on $a$ and consider weight functions
  $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and $W_b\colon 2^{\{a, b\}} \to
  \mathbb{R}_{\ge 0}$ defined as
  \begin{align*}
    W_a(\{a\}) &= 0.9, \\
    W_a(\emptyset) &= 0.1,
  \end{align*}
  and
  \begin{align*}
    W_b(\{a, b\}) &= 0.8, \\
    W_b(\{a\}) &= 0.2, \\
    W_b(\{b\}) &= 0.7, \\
    W_b(\emptyset) &= 0.3.
  \end{align*}
  One can easily check that with these definitions $W$ indeed induces $M$.
\end{example}

@@

\begin{theorem}
  A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is factorable if and only
  if
  \begin{equation} \label{eq:wmccondition}
  M(u \land v) = M(u)M(v)
  \end{equation}
  for all distinct $u, v \in U \cup \{ \neg w \mid w \in U \}$ such that $u \ne
  \neg v$.
\end{theorem}
\begin{proof}
  ($\Leftarrow$) For each $x \in U$, let $W_x\colon 2^{\{x\}} \to
  \mathbb{R}_{\ge 0}$ be defined by $W_x(\{ x \}) = M(x)$ and $W_x(\emptyset) =
  M(\neg x)$. Let $M_W$ be the measure induced by $W = \prod_{x \in U} W_x$. We
  will show that $M = M_W$. First, note that $M_w(\bot) = 0 = M(\bot)$ by
  \cref{def:wmc,prop:measure}. Second, let $a = \bigwedge_{u \in U} a_u$ be an
  atom in $2^{2^U}$ such that $a_u \in \{ u, \neg u \}$ for all $u \in U$. Then
  \[
    M_W(a) = W(\{ a_u \mid u \in U \}) = \prod_{u \in U} W_u(\{a_u\}) = \prod_{u
      \in U} M(a_u) = M \left(\bigwedge_{u \in U} a_u \right) = M(a)
  \]

@@

  ($\Rightarrow$) For the other direction, we are given a factored weight
  function $W = \prod_{x \in U} W_x$, and we want to show that its induced
  measure $M_W$ satisfies \cref{eq:wmccondition}. Let $k_u, k_v \in U \cup \{
  \neg w \mid w \in U \}$ be such that $k_u \in \{ u, \neg u \}$, $k_v \in \{ v,
  \neg v \}$, and $u \ne v$. We then want to show that
  \begin{equation} \label{eq:to_prove3}
    M_W(k_u \land k_v) = M_W(k_u)M_W(k_v)
  \end{equation}
  which is equivalent to
  \begin{equation} \label{eq:to_prove4}
    M_W(k_u \land k_v) \cdot M_W(\neg k_u \land \neg k_v) = M_W(k_u \land \neg k_v) \cdot M_w(\neg k_u \land k_v)
  \end{equation}
  by \cref{lemma:before_theorem}. Then
  \begin{align*}
    M_W(k_u \land k_v) &= \sum_{\{a\} \le k_u \land k_v} W(a) = \sum_{\{a\} \le k_u \land k_v} \prod_{x \in U} W_x(a) \\
                        &= \sum_{\{a\} \le k_u \land k_v} W_u(a_u)W_v(a_v) \prod_{x \in U \setminus \{ u, v \}} W_x(a) = \sum_{\{a\} \le k_u \land k_v} W_u(k_u)W_v(k_v) \prod_{x \in U \setminus \{ u, v \}} W_x(a) \\
    &= W_u(k_u)W_v(k_v) \sum_{\{a\} \le k_u \land k_v} \prod_{x \in U \setminus \{ u, v \}} W_x(a) = W_u(k_u)W_v(k_v)C,
  \end{align*}
  where $C = \sum_{\{a\} \le k_u \land k_v} \prod_{x \in U \setminus \{ u, v \}}
  W_x(a)$, and is the same for $M_W(\neg k_u \land k_v)$, $M_W(k_u \land \neg
  k_v)$, and $M_W(\neg k_u \land \neg k_v)$ as well. But then
  \cref{eq:to_prove4} becomes
  \[
    W_u(k_u)W_v(k_v)W_u(\neg k_u)W_v(\neg k_v)C^2 = W_u(k_u)W_v(\neg
    k_v)W_u(\neg k_u)W_v(k_v)C^2
  \]
  which is trivially true.
\end{proof}

@@

A \emph{Bayesian network} is a directed acyclic graph with random variables as
vertices that defines a probability distribution over them. Let $\mathcal{V}$
denote the set of random variables and $X \in \mathcal{V}$ be a random variable.
Let $\im X$ denote the set of possible values of $X$ and $\mathrm{pa}(X)$ denote
the set of its parents. The full probability distribution is then equal to
\begin{equation} \label{eq:factorisation}
  \prod_{X \in \mathcal{V}} \Pr(X \mid \mathrm{pa}(X)).
\end{equation}
If all random variables are discrete (and we only consider discrete Bayesian
networks in this paper), the probabilities contained in each factor of
\cref{eq:factorisation} can be summarised in a table known as a
\emph{conditional probability table} (CPT). See \cref{fig:example_bn} for an
example Bayesian network which we will refer to throughout this section. For
this network, $\mathcal{V} = \{ W, F, T \}$, $\mathrm{pa}(W) = \emptyset$,
$\mathrm{pa}(F) = \mathrm{pa}(T) = \{ W \}$, $\im W = \im F = \{0, 1 \}$, and
$\im T = \{ l, m, h \}$.

\begin{definition}[Indicator variables]
  Let $X \in \mathcal{V}$ be a random variable. If $X$ is binary (i.e., $|\im X|
  = 2$), we can arbitrary identify one of the values as $1$ and the other one as
  $0$ (i.e, $\im X \cong \{ 0, 1 \}$). Then $X$ can be represented by a single
  \emph{indicator variable} $\lambda_{X=1}$. For notational simplicity, for any
  set $S$, we write $\lambda_{X=0} \in S$ or $S = \{ \lambda_{X=0}, \dots \}$ to
  mean $\lambda_{X=1} \not\in S$.

  On the other hand, if $X$ is not binary, we represent $X$ with $|\im X|$
  indicator variables, one for each value. We let
  \[
    E(X) = \begin{cases}
      \{ \lambda_{X=1} \} & \text{if } |\im X| = 2 \\
      \{ \lambda_{X=x} \mid x \in \im X \} & \text{otherwise.}
    \end{cases}
  \]
  denote the set of indicator variables for $X$ and $E^*(X) = E(X) \cup
  \bigcup_{Y \in \mathrm{pa}(X)} E(Y)$ denote the set of indicator variables for
  $X$ and its parents in the Bayesian network. Finally, let $U = \bigcup_{X \in
    \mathcal{V}} E(X)$ denote the set of all indicator variables for all random
  variables in the Bayesian network.
\end{definition}

@@

\begin{algorithm}
  \caption{Encoding a Bayesian network as a function $2^U \to \mathbb{R}_{\ge
      0}$}
  \label{alg:encoding}
  \KwData{a Bayesian network with vertices $\mathcal{V}$ and probability
    distribution $\Pr$}
  \KwResult{a function $\phi\colon 2^U \to \mathbb{R}_{\ge 0}$}
  $\phi \gets 1$\;
  \For{$X \in \mathcal{V}$}{
    \textit{let} $\mathrm{pa}(X) = \{ Y_1, \dots, Y_n \}$\;
    $\mathrm{CPT}_X \gets 0$\;
    \eIf{$|\im X| = 2$}{
      \For{$(y_1, \dots, y_n) \in \prod_{i = 1}^n \im Y_i$}{
        $p_1 \gets \Pr(X = 1 \mid Y_1 = y_1, \dots, Y_n = y_n)$\;
        $p_0 \gets \Pr(X \ne 1 \mid Y_1 = y_1, \dots, Y_n = y_n)$\;
        $\mathrm{CPT}_X \gets \mathrm{CPT}_X + p_1[\lambda_{X=1}] \cdot
        \prod_{i=1}^n [\lambda_{Y_i=y_i}] + p_0 \overline{[\lambda_{X=1}]} \cdot
        \prod_{i=1}^n [\lambda_{Y_i=y_i}]$\;
      }
    }{
      \textit{let} $\im X = \{ x_1, \dots, x_m \}$\;
      \For{$x \in \im X$ {\rm \textbf{and}} $(y_1, \dots, y_n) \in \prod_{i =
          1}^n \im Y_i$}{
        $p_x \gets \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n)$\;
        $\mathrm{CPT}_X \gets \mathrm{CPT}_X + p_x[\lambda_{X=x}] \cdot
        \prod_{i=1}^n [\lambda_{Y_i=y_i}] + \overline{[\lambda_{X=1}]} \cdot
        \prod_{i=1}^n [\lambda_{Y_i=y_i}]$\;
      }
      $\mathrm{CPT}_X \gets \mathrm{CPT}_X \cdot \left( \sum_{i=1}^m [\lambda_{X
          = x_i}] \right) \cdot \prod_{i=1}^m \prod_{j=i+1}^m
      (\overline{[\lambda_{X = x_i}]} + \overline{[\lambda_{X = x_j}]})$\;
    }
    $\phi \gets \phi \cdot \mathrm{CPT}_X$\;
  }
  \Return{$\phi$}\;
\end{algorithm}

@@

\begin{lemma} \label{lemma:cpt}
  Let $X \in \mathcal{V}$ be a random variable with parents $\mathrm{pa}(X) = \{ Y_1,
  \dots, Y_n \}$. Then $\mathrm{CPT}_X\colon 2^{E^*(X)} \to \mathbb{R}_{\ge 0}$
  is such that for any $x \in \im X$ and $(y_1, \dots, y_n) \in \prod_{i=1}^n
  \im Y_i$,
  \[
    \mathrm{CPT}_X (\{ \lambda_{X=x} \} \cup \{ \lambda_{Y_i=y_i} \mid i = 1,
    \dots, n \}) = \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n).
  \]
\end{lemma}
\begin{proof}
  Let $\tau = \{ \lambda_{X=x} \} \cup \{ \lambda_{Y_i=y_i} \mid i = 1, \dots, n
  \}$. If $X$ is binary, then $\mathrm{CPT}_X$ is a sum of $2\prod_{i=1}^n |\im
  Y_i|$ terms, one for each possible assignment of values to variables $X, Y_1,
  \dots, Y_n$. Exactly one of these terms is nonzero when applied to $\tau$, and
  it is equal to $\Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n)$ by definition.

  If $X$ is not binary, then $\left( \sum_{i=1}^m [\lambda_{X = x_i}]
  \right)(\tau) = 1$, and
  \[
    \left( \prod_{i=1}^m \prod_{j=i+1}^m (\overline{[\lambda_{X = x_i}]} +
      \overline{[\lambda_{X = x_j}]}) \right)(\tau) = 1,
  \]
  so $\mathrm{CPT}_X(\tau) = \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n)$ by a
  similar argument as before.
\end{proof}
